Self-Supervised Transfer Learning From 2D Images to 3D Point Clouds

### 2022-05-10 doc/paper.tex
adds related works

\textbf{Sparse 3D convolutions.} Talk about submanifold networks, Minkowski Engine \cite{choy20194d}, pointnet \cite{charles2017PointNet} PointNet++ \cite{qi2017pointnet++}.

\textbf{Downstream tasks.} PointGroup \cite{}

\textbf{Contrastive Learning.} Talk about SimCLR \cite{chen_simple_2020}, BOYL \cite{grill_bootstrap_nodate}, Moco \cite{he_momentum_2020}. Then talk about PointContrast. \cite{choy2019Fully}

DenseCL \cite{wang_dense_2021} Does contrastive learning at the pixel level of images giving a good background into extracting image level features.

\textbf{Contrastive learning variations}
Supervised contrastive losses \cite{khosla_supervised_2021} Guided contrastive learning in 3D \cite{jiang2021Guided}. Prototypical learning \cite{li_prototypical_2021}

\textbf{Contrastive Learning in 3D}
DepthContrast \cite{zhang2021Self}, PointContrast \cite{xie2020pointcontrast}, Contrastive scene contexts \cite{hou2021Exploring}

Occlusion completion \cite{wang_unsupervised_nodate}

\textbf{Using 2D images.} 3D-sis \cite{hou20193d}, Using 2D images to learn 3D segmentation \cite{genova2021learning} \cite{frey_continual_2021}. 3dmv \cite{angela2018joint}

Closest to my work: \cite{yunze2020p4contrast}. Their approach is to augment 3D points with features from 2D images and use their combined multi-modal features as the output of their model. They require however that their model always have access to 2D images and cannot operate on solely colorized point clouds at runtime.

@@

The goal of self-supervised transfer learning is to learn from unlabeled data using a self-supervised objective which in turn initializes a model's backbone. The initialization is meant to improve performance of downstream tasks. Specifically, contrastive learning uses an objective function that encourages the outputs of a model on similar datapoints to be close together while keeping those of dissimilar datapoints to be far apart, in some metric space.

@@

\AJ{Add figure of what your feature extractor looks like}

\AJ{Need to review how I do this in the code, I think I use a projector left over from BYOL}

@@

\AJ{Do I need to give the equation for Bi-linear interpolation or is this also considered basic knowledge}

@@

\AJ{Need to look at not using a projection head in 3D model.}

\clearpage
\section{Experiments}
\label{sec:results}

@@

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=8cm]{images/plots/s3dis_semantic.pdf}
    \includegraphics[width=8cm]{images/plots/scannet_0.05_semantic.pdf}
    \includegraphics[width=8cm]{images/plots/scannet_semantic.pdf}
    \caption{Comparison of validation performance over training steps for models pre-trained on Scannet and fine-tuned for semantic segmentation on: \textbf{Top:} S3DIS, \textbf{Middle:} Scannet with only 5\% labeled data, \textbf{Bottom:} Scannet}
    \label{fig:validationComparison}
\end{figure}

>>>>

Our approach builds upon work on feature extraction and self-supervised learning in images and point clouds. Here we provide a brief summary of work in these fields.

\textbf{Feature Extraction in 3D.} Raw sensor measurements capture 3D scenes as either depth or points. These points require processing for downstream tasks such as semantic segmentation or object detection. This can be accomplished in two ways. The first is to use the raw points themselves \cite{charles2017PointNet, qi2017pointnet++, Tatarchenko2018TangentConvolutions, thomas2019KPConv, wu2019PointConv, engelmann2020Dilated, zhao2020point} which does not require any other expensive transformations. The other is to transform these points into an intermediate representation, such as discrete voxels \cite{cicek20163d, riegler2017octnet, graham20183D, choy20194d}. Any of these approaches can equally be used for extracting meaningful features for downstream tasks. However, in this work we chose to use the 3D U-Net voxel-based convolution presented in Chow \etal \cite{choy20194d}. Not only does it offer close to state of the art performance but many recent state of the art methods \cite{jiang2020pointgroup, chen2021hierarchical} have adopted this voxel-based feature extractor.

\textbf{Self-Supervised Transfer learning.} Self-supervised learning uses a pretext task that operates on raw data directly to form a loss function that learns meaningful representations. It can take the form of generative models such as auto-encoders \cite{goodfellow2014generative, Kingma2014}, which learn how to model the input and make examples that are close to the real distribution. Or discriminative models, that aim to discriminative between inputs in high dimensional space.

Contrastive learning is a discriminative learning approach which contrasts different inputs to learn how to discriminate between positive and negative samples. It has been shown to be incredibly powerful in image classification. SimCLR \cite{chen_simple_2020}  uses a very simple framework which achieves close to supervised level performance. Some methods focus on increasing the batch size through a memory bank \cite{wu2018unsupervied} or momentum encoder \cite{he_momentum_2020}.

Choosing which points to contrast is a huge challenge as some points are uninformative and others may belong to the same class. To address this issue, some methods select samples in a more informative way using negative mining techniques \cite{robinson_contrastive_2021, kalantidis_hard_nodate, chuang_debiased_2020, ho_contrastive_nodate} to improve the strength of the self-supervised signal. Other methods try to bootstrap which points should be contrasted. Li \etal \cite{li2021prototypical} iteratively groups representations into prototypes to avoid contrasting points from the same cluster. Khosla \etal \cite{khosla_supervised_2021} use a mix of supervised learning to generate pseudo labels that inform the selection of points during self-supervised learning.

The bulk of this work has been on high-level image classification, but it has proved useful in pixel level image segmentation as well \cite{wang_dense_2021, xie2021propagate, wang2021exploring}. Since these features were useful in 2D segmentation, it suggests that they can be equally as useful in 3D.

\textbf{Contrastive Learning in 3D}

The same techniques that were used in images can translate directly into 3D point clouds. This was first used for generating unique features for point clouds registration \cite{choy2019Fully}. This was subsequently adapted by PointContrast \cite{xie2020pointcontrast} to work for semantic segmentation. Point pairs are generated using two scans of registered overlapping point clouds and finding points which are spatially close in both. The scans are randomly transformed to learn transformation invariant features. Contrastive scene contexts \cite{hou2021Exploring} expands on this idea by sampling points from partitioned spatial contexts. This work extends point contrast to use instead of a second scan, to use features from an image of the same scan.

Alternatively, DepthContrast \cite{zhang2021Self} seeks to learn point-level representations using scan level features. They implement a momentum encoder \cite{he_momentum_2020} to operate on scene level descriptors generated from individual scans which are then contrasted with each other. Using other high level tasks like occlusion completion \cite{wang_unsupervised_nodate} has also proven useful.

The same sampling challenges exist in point clouds as they do for images. To handle this \cite{jiang2021Guided} use pseudo labels to pick points to avoid contrasting from the same class.

\textbf{Sensor Fusion.} Our work seeks to leverage features learned exclusively from images and transfer them to a 3D model. Combining the two modalities has proven successful in learning 3D segmentation \cite{hou20193d, genova2021learning, frey_continual_2021, angela2018joint} as well as using multiple images of a single scene in multi-view fusion \cite{kundu2020virtual}.

P4Contrast \cite{yunze2020p4contrast} combines sensor fusion with self-supervised pre-training. Their approach is to augment 3D points with features from 2D images and use their combined multi-modal features as the output of their model. Their approach is to fuse both modalities during pre-training and runtime and so always require access to 2D images. This means they cannot operate on solely colorized point clouds at runtime. Our method seeks to only take learned 2D representations and transfer them to a 3D model during pre-training.

@@

This section will layout a theoretical framework for contrastive learning that will be used throughout the rest of this work.
The goal of self-supervised transfer learning is to learn from unlabeled data using a self-supervised objective which in turn initializes a model's backbone. The initialization is meant to improve performance of downstream tasks. Specifically, contrastive learning uses an objective function that encourages the outputs of a model on similar datapoints to be close together while keeping those of dissimilar datapoints to be far apart, in some metric space.

@@

@@

@@

\section{Experiments}
\label{sec:results}

@@

\begin{figure}
    \centering
    \includegraphics[width=8cm]{images/plots/s3dis_semantic.pdf}
    \includegraphics[width=8cm]{images/plots/scannet_0.05_semantic.pdf}
    % \includegraphics[width=8cm]{images/plots/scannet_semantic.pdf}
    \caption{Comparison of validation performance over training steps for models pre-trained on Scannet and fine-tuned for semantic segmentation on: \textbf{Top:} S3DIS, \textbf{Bottom:} Scannet with only 5\% labeled data}
    \label{fig:validationComparison}
\end{figure}

@@
